The Power of Two Random Choices and the Surprising Influence of Asymmetry\\ \normalsize A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Tr\"aff and Francesco Versaci, ``Parallel Algorithms'' lecture, 2012 winter term at TU Wien.}

### 2014-04-09 article/article.tex
use symlinks for source files
some minor fixes to finalize the paper

\title{The Power of Two Random Choices\\ 
\large A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Tr\"aff and Francesco Versaci, ``Parallel Algorithms'' lecture, 2012 winter term at TU Wien.}}
\author{Martin Kalany, 0825673}
\maketitle

\begin{abstract}
Balls-into-bins is a classical allocation problem, where $m$ balls have to be distributed among $n$ bins as evenly as possible. A well studied approach is to select a bin uniformly and independently at random for each ball. Another strategy is to select $d\geq 2$ bins for each ball and put the ball into the bin with the least load. Somewhat surprisingly, the maximum load achieved by such a multiple-choice scheme is significantly smaller, even for $d=2$. In this paper, we will present two different multiple-choice allocation strategies for the balls-into-bins problem and provide a detailed analysis of the upper bounds on the maximum load that they achieve. We will use a technique called witness trees, which has proven useful in a wide variety of balls-into-bins problems.
\end{abstract}

\section{Introduction}
\label{sec:intro}
Suppose that $m$ balls are to be thrown into $n$ bins, according to some placement strategy. We are interested in the number of balls in the fullest bin, also called the \emph{maximum load}. Equivalently, one can look at the \emph{additive gap} of the allocation, i.e., the difference of the number of balls in the fullest bin and the average amount of balls in the bins. These balls-into-bins games (BiB) have been studied extensively (see, e.g.,~\cite{JK77}). In the context of computer science, the BiB model is useful to study, e.g., strategies for distributing tasks (balls) to processors (bins) as evenly as possible without having to query all the processors for the number of tasks currently assigned to them.

While most of this survey and the analyses focuses on the lightly loaded case where $m=n$, we will give results for the case $m > n$ as well. We will denote the maximum load by $\load$ and the additive gap by $\delta$:
\begin{align*}
\delta = \load - \frac{m}{n} \quad .
\end{align*}

@@

\subsection{Multiple-choice BiB}
Consider a variation of the  single-choice BiB strategy: For each ball, choose $d \geq 2$ bins independently and uniformly at random and place the ball into the least full bin. If there are multiple bins with the same minimal load, choose one among them arbitrarily. This simple multiple-choice scheme, called the \emph{Greedy} algorithm, achieves a maximum load of
\begin{align*}
\load = \frac{m}{n}+\frac{\ln \ln n}{\ln d} + O(1) \implies \delta = \frac{\ln \ln n}{\ln d} + O(1) \quad ,
\end{align*}
w.h.p., for all $m$~\cite{ABKU99, BCSV06}, i.e., this additional choice leads to an exponential reduction of the additional gap. 
\begin{comment}
Curiously, the exponential decrease in the gap is already achieved having just two choices (that is, $d=2$). Each additional choice decreases the bound for the maximum gap by just a constant factor~\cite{MRS01}. 
\end{comment}
Extensive research following the influential paper by Azar et al.~\cite{ABKU99}, who were the first ones to explicitly prove this phenomenon\footnote{For a brief history of the proof, which is quite interesting by itself, refer to Appendix \ref{sec:historyOfGreedy}.}, suggests that this \emph{two-choice paradigm} is not just an artefact of the simple BiB model, but a general principle applicable to a variety of problems.

Consider the slightly more involved \emph{Always-go-left} scheme: Suppose that the $n$ bins are partitioned evenly into $d$ groups, each of size $\lfloor \frac{n}{d}\rfloor$ or $\left\lceil\frac{n}{d}\right\rceil$ (to simplify the presentation, we will assume that $d$ divides $n$ evenly in the remainder of this survey). We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. For tie-breaking, we assume a fixed ordering of bins $i$ for $1 \leq i \leq n$ and put the ball into the bin with smallest index $i$ among those with the same minimal load. Somewhat surprisingly, V\"ocking~\cite{VOC03} showed that such a strategy employing asymmetric tie-breaking in combination with non-uniform bin selection reduces the bound for the maximum load to
\begin{align*}
\load = \frac{m}{n} + \frac{\ln\ln n}{d  \ln \Phi_d} + O(1),
\end{align*}
w.h.p., for all $m$. (The symbol $\Phi_d$ denotes a generalization of the golden ratio, based on generalized Fibonacci numbers. Refer to Appendix \ref{sec:fibonacci} for the definition.) Berenbrink et al.~\cite{BCSV06} and V\"ocking~\cite{VOC03} were able to show that the Always-go-left strategy is optimal up to additive constants.

\subsubsection*{Outline}
In Section \ref{sec:classification}, we give a basic classification of BiB games to provide an overview of the research, followed by some applications in Section \ref{sec:applications}. Section \ref{sec:model} contains a detailed definition of the model we use, which can be classified as a sequential, finite and dynamic process with random deletions. In Section \ref{sec:multiple-choice} we provide a brief definition of three different classes of random placement strategies and present two placement algorithms outlined in the introduction: Greedy (Section \ref{sec:greedy}) and Always-go-left (Section \ref{sec:AlwaysGoLeft}). Section \ref{sec:analysis} contains a detailed analysis of both strategies for the case $m=n$.

\section{Classification of BiB games}
\label{sec:classification}
Several different variations of this problem exist, which may be classified as follows:
\begin{compactitem}
\item In a \emph{sequential} process, the balls arrive one after another and a ball is dealt with only after the previous one has been placed. Conversely, in a \emph{parallel} process, balls arrive in sets and the balls in one set are dealt with simultaneously~\cite{ABS98}. 
\item In a \emph{static} problem, the balls are placed into the bins and no balls are ever removed, whereas in a \emph{dynamic} problem balls may also be removed from the system. In the random deletions model, for each newly inserted ball another one is chosen uniformly and independently at random and removed from the system~\cite{ABKU99, MRS01}. More elaborate schemes exists: An oblivious adversary might specify a sequence of insertions and deletions of balls in advance, without any knowledge of their placement~\cite{CFM+98}. Adler et al.~\cite{ABS98} studied another approach in the context of the parallel BiB problem, where the balls are stored in FIFO-order and at every time step the first ball in each bin is deleted. 
\item A \emph{finite} process is analysed for a given time interval that is known beforehand, whereas for an \emph{infinite} one, placements and deletions are given for an infinite time span~\cite{ABS98}.
\item In the lightly loaded case, $m$ is in the order of $n$, whereas in the heavily loaded case, $m$ is arbitrarily large. W.l.o.g., we will assume that $m = n$ for the lightly loaded case. 
\end{compactitem}

@@

\section{BiB model}
\label{sec:model}
Although several different variants of the BiB problem exist, we will focus on a basic model to highlight the crucial ideas in the placement strategies and their analyses. They are based on a sequential, infinite and dynamic process that uses an oblivious adversary, which provides a sequence $\sigma = \sigma_1 \sigma_2 \dots$, where $\sigma_t = (d_t, a_t)$ represents a request for a ball $d_t$ to be deleted and a different ball $a_t$ to be added to the bins at time step $t$. In our model, the ball $d_t$ associated with some request $\sigma_t$ is deleted at the beginning of time step $t$, next, the ball $a_t$ has to be placed into one of the $n$ bins, without any knowledge of the following requests $\sigma_{t+1}$, $\sigma_{t+2},\dots$ This infinite sequence of insertions and deletions is fixed in advance and is thus independent of previously allocated balls and any decisions made by the allocation algorithm. We assume that at any time $t$ at most $m = n$ balls are in the bins.

\section{Multiple-choice BiB}
\label{sec:multiple-choice}
The single-choice BiB strategy (Section \ref{sec:single-choice}) chooses one bin uniformly and independently at random for each ball. A multiple-choice BiB strategy generalizes this idea by  choosing $d \geq 2$ bins and placing the ball into the least full. We distinguish placement algorithms depending on how the $d$ locations (i.e, bins) are sampled~\cite{VOC03}.

@@

\subsection{The Greedy scheme}
\label{sec:greedy}
Consider the Greedy scheme: For each ball, choose $d \geq 2$ bins independently and uniformly at random and place the ball into the least full bin. If there are multiple bins with the same minimal load, chose one among them arbitrarily. Note that the $d$ selected bins are not necessarily distinct.

\begin{theorem}
\label{theorem:greedy}
Consider any sequence of insertions and deletions s.t.~at most $n$ balls are in the bins at any time. If the balls are placed into bins by the Greedy algorithm, then the maximum load is at most
\begin{align*}
\load \leq \frac{\ln \ln n}{\ln d}+ O(1) \quad ,
\end{align*}
w.h.p., for all $t$~\cite{ABKU99, BCSV06}.
\end{theorem}
This result is remarkable in several ways:
\begin{compactitem}
\item The apparently small change made for the Greedy algorithm compared to the single-choice BiB scheme results in an exponential decrease of the gap, even for $d=2$. 
\item Each additional choice ($d > 2$) decreases the gap by only a constant factor~\cite{MRS01}.
\end{compactitem}

@@

\begin{theorem}
\label{theorem:aglm}
Consider any sequence of insertions and deletions s.t.~at most $n$ balls are in the bins at any time. If the balls are placed into bins by the Always-go-left algorithm, the maximum load is at most 
\begin{align*}
\load \leq \frac{\ln\ln n}{d  \ln \Phi_d} + O(1) \quad ,
\end{align*}
w.h.p.,~for all $t$~\cite{VOC03}.
\end{theorem}
\noindent
V\"ocking~\cite{VOC03} also proved a matching lower bound for the lightly loaded case:
\begin{theorem}
\label{theorem:aglmLower}
For $m=n$, using an arbitrary allocation scheme that chooses $d$ bins for each ball at random (i.e., any allocation scheme in class 3) the maximum load is
\begin{align*}
\load = \frac{\ln \ln n}{d \ln \Phi_d} - O(1) \quad 
\end{align*}
\end{theorem}
\noindent
w.h.p.,~for all $t$~\cite{VOC03}. We not that the combination of asymmetric tie-breaking and and non-uniform (but independent) sampling of the $d$ bins is crucial to achieve bounds given in Theorems~\ref{theorem:aglm} and~\ref{theorem:aglmLower}.

@@

We will prove the bounds for $m=n$ for both the Greedy and the Always-go-left scheme in Sections~\ref{sec:analysis:greedy} and~\ref{sec:analysis:alg}, respectively. However, we will adopt a simplifying assumption to provide a more clear presentation of the proof technique. Section~\ref{sec:analysis:nondistinctBalls} will extend these proofs to remove this assumption.

We will use a proof technique called witness trees, which allows for a clean analysis of both problems and is applicable to a wide variety of BiB scenarios. Other proof techniques such as layered induction and a differential equation approaches have been used in the literature; see, e.g.,~\cite{MRS01} for an overview of different proof techniques for the BiB problem. The proof for arbitrarily large $m$ is more involved and we refer the interested reader to~\cite{BCSV06, TW13}.

@@

\subsubsection{Symmetric witness trees}
\label{sec:analysis:definitionSymWT}
A symmetric witness tree (SWT) of order $L$ is a complete $d$-ary tree with $d^{L}$ leaves, where each node $v$ references a ball $b_v = \mathrm{ball}(v)$ and the nodes and edges of the SWT represent events that may occur or not. In general, the same ball may be referenced by several nodes. However, initially we will make use of the following assumption in the analysis.
\begin{assumption}
\label{assumption:independence}
All the events represented by the nodes and edges of a witness tree are stochastically independent. 
\end{assumption} 
\noindent
This implies that no ball is referenced by more than one node of a witness tree. A correct analysis will have to consider balls assigned to more than one node, but this simplifying assumption is useful to highlight the differences between the symmetric and asymmetric allocation schemes. Section~\ref{sec:analysis:nondistinctBalls} will extend the analysis to witness trees with redundant balls.

We define edge and leaf events:
\begin{description}
\item [Edge event:] Consider some edge $e = (u,v)$ of the witness tree, s.t.~$v$ is the $i$-th child of $u$. The edge $e$ represents the event that the $i$-th location of the ball $b_u$ points to the same bin as one of the locations of the ball $b_v$. 
\item [Leaf event:] A leaf node $v$ of the witness tree represents the event that each of the $d$ locations of the ball $b_v$ points to a bin that contains at least three additional balls (i.e., balls that are not represented by a node of the tree), at the time of the insertion of the ball $b_v$.
\end{description}
To avoid dependencies, the crucial step is to define edge events in terms of the possible locations of balls, not by their finally assigned bin.

@@

\subsubsection{Probability of activation}
\label{sec:analysis:probabilitySymWT}
In this section, we derive the probability that some SWT of order $L$ is activated, which gives an upper bound for the probability that the maximum load is greater than $L+3$ at some time $t$.

@@

The probability that a leaf event is activated is bounded from above by $3^{-d}$: Recall that by the definition of leaf events, each of the $d$ locations of a ball $b_v$, where $v$ is a leaf node, has to point to a bin with at least 3 additional balls in it. The number of those bins is bounded by $\lfloor n/3 \rfloor$ at all time steps $t$, implying a probability of at most $1/3$ that one such bin is chosen. This implies a probability of at most $3^{-d}$ that all the $d$ bins are chosen from among those with at least 3 balls in them. Thus, the probability that all leaf events are activated is at most $p_\mathrm{l} = 3^{-d l}$, where $l = d^{L}$ is the number of leaves in an SWT.

Thus, the probability that an activated SWT exists at time $t$ is at most 
\begin{align*}
P_\mathrm{s} &\leq w  p_\mathrm{e}  p_\mathrm{l} 
		= n^{k}  \left(\frac{d}{n}\right)^{k-1}  3^{-d  l} 
		= n  d^{k-1}  3^{-d  l}\quad .
\end{align*}
Using $k \leq 2l$ we get
\begin{align*}
P_\mathrm{s} &\leq n  d^{2 l}  3^{-d  l} 
		= n  \left(d^2\right)^l  \left( 3^{-d} \right)^l\quad ,
\end{align*}
which we simplify by using $2d^2 \leq 3^d$ as
\begin{align*}
P_\mathrm{s} &\leq n  \left(\frac{3^d}{2}\right)^l  \left( 3^{-d}\right)^l 
		= n  2^{-l}  3^{d l}  3^{-d  l} 
		= n 2^{-l}		
		= n 2^{-d^L} \quad .		
\end{align*}
For any $\alpha > 0 $, let
\begin{align*}
L &\geq \log_d\log_2 n + \log_d\left(1+\alpha\right) \neqn{swt:chooseL} \quad .
\end{align*}
Then, 
\begin{align*}
P_\mathrm{s} &\leq n  2^{{-d}^{\log_d\log_2 n + \log_d\left(1+\alpha\right)}} 
		= n  2^{-{d}^{\log_d\log_2 n} d^{\log_d \left(1+\alpha\right)}} \\
		&= n  \left( 2^{-\log_2 n  \left(1+ \alpha \right)} \right) 
		= n  n^{-\left(1+\alpha\right)} = n^{-\alpha} \neqn{swt:result}\quad .
\end{align*}
Thus, for any $\alpha>0$ and large enough $L$ as specified in Eq.~\ref{eqn:swt:chooseL} the probability that an SWT of order $L$ is activated is bounded from above as $P_\mathrm{s}\leq n^{-\alpha}$.
\end{proof}

Since the probability for the activation of an SWT of order $L$ upper-bounds the probability of the bad event that some bin contains more than $L+3$ balls, Eq.~\ref{eqn:swt:result} is also an upper bound for the latter. Note that Eq.~\ref{eqn:swt:chooseL} can be rewritten as
\begin{align*}
L &\geq  \frac{\ln\log_2 n}{\ln d} + \frac{\ln(1+\alpha)}{\ln d} 
   = \frac{\ln \log_2 n}{\ln d} +O(1) 
  \geq \frac{\ln \ln n}{\ln d} +O(1) \neqn{swt:boundL} \quad ,
\end{align*}
thus proving Theorem~\ref{theorem:greedy}, under Assumption~\ref{assumption:independence} (i.e., that all events are stochastically independent). \qed

@@

\subsubsection{Asymmetric witness trees}
\label{sec:analysis:definitionAsymWT}
An AWT differes from an SWT (Section~\ref{sec:analysis:definitionSymWT}) only by its structure, which is a $d$-step Fibonacci tree $T_d\left(k \right)$:

>>>>

\title{The Power of Two Random Choices and the Surprising Influence of Asymmetry\\ 
\normalsize A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Tr\"aff and Francesco Versaci, ``Parallel Algorithms'' lecture, 2012 winter term at TU Wien.}}
\author{Martin Kalany, 0825673}
\maketitle

\begin{abstract}
\noindent
Balls-into-bins is a classical allocation problem, where $m$ balls have to be distributed among $n$ bins as evenly as possible. A well studied approach is to select a bin uniformly and independently at random for each ball; another strategy is to select $d\geq 2$ bins for each ball and put the ball into the bin with the least load. Somewhat surprisingly, the maximum load achieved by such a multiple-choice scheme is significantly smaller, even for $d=2$. In this paper, we will present two different multiple-choice allocation strategies for the balls-into-bins problem and provide a detailed analysis of the upper bounds on the maximum load that they achieve. We will use a technique called witness trees, which has proven useful in a wide variety of balls-into-bins problems.
\end{abstract}

\section{Introduction}
\label{sec:intro}
Suppose that $m$ balls are to be thrown into $n$ bins, according to some placement strategy. We are interested in the number of balls in the fullest bin, also called the \emph{maximum load}. Equivalently, we can look at the \emph{additive gap} of the allocation, i.e., the difference of the number of balls in the fullest bin and the average amount of balls in the bins. These balls-into-bins games (BiB) have been studied extensively (see, e.g.,~\cite{JK77}). In the context of computer science, the BiB model is useful to study, e.g., strategies for distributing tasks (balls) to processors (bins) as evenly as possible without having to query all the processors for the number of tasks currently assigned to them.

While most of this survey and the analyses focuses on the lightly loaded case where $m=n$, we will give results for the case $m > n$ as well. We denote the maximum load by $\load$ and the additive gap by $\delta$:
\begin{align*}
\delta = \load - \frac{m}{n} \quad .
\end{align*}

@@

\subsection{Multiple-choice BiB}
Consider the following variation of the  single-choice BiB strategy: For each ball, choose $d \geq 2$ bins independently and uniformly at random and place the ball into the least full bin. If there are multiple bins with the same minimal load, choose one among them arbitrarily. This simple multiple-choice scheme, called the \emph{Greedy} algorithm, achieves a maximum load of
\begin{align*}
\load = \frac{m}{n}+\frac{\ln \ln n}{\ln d} + O(1) \implies \delta = \frac{\ln \ln n}{\ln d} + O(1) \quad ,
\end{align*}
w.h.p., for all $m$~\cite{ABKU99, BCSV06}, i.e., this additional choice leads to an exponential reduction of the additive gap. 
\begin{comment}
Curiously, the exponential decrease in the gap is already achieved having just two choices (that is, $d=2$). Each additional choice decreases the bound for the maximum gap by just a constant factor~\cite{MRS01}. 
\end{comment}
Extensive research following the influential paper by Azar et al.~\cite{ABKU99}, who were the first ones to explicitly prove this phenomenon\footnote{For a brief history of the proof, which is quite interesting by itself, refer to Appendix \ref{sec:historyOfGreedy}.}, suggests that this \emph{two-choice paradigm} is not just an artefact of the simple BiB model, but a general principle applicable to a variety of problems.

Consider the slightly more involved \emph{Always-go-left} scheme: Suppose that the $n$ bins are partitioned evenly into $d$ groups, each of size $\lfloor \frac{n}{d}\rfloor$ or $\left\lceil\frac{n}{d}\right\rceil$ (to simplify the presentation, we will assume that $d$ divides $n$ evenly in the remainder of this survey). We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. For tie-breaking, we assume a fixed ordering of bins $i$ for $1 \leq i \leq n$ and put the ball into the bin with smallest index $i$ among those with the same minimal load. Somewhat surprisingly, V\"ocking~\cite{VOC03} showed that such a strategy employing asymmetric tie-breaking in combination with non-uniform bin selection reduces the bound for the maximum load to
\begin{align*}
\load = \frac{m}{n} + \frac{\ln\ln n}{d  \ln \Phi_d} + O(1),
\end{align*}
w.h.p., for all $m$. (The symbol $\Phi_d$ denotes a generalization of the golden ratio, based on generalized Fibonacci numbers. Refer to Appendix \ref{sec:fibonacci} for the definition.)

\subsubsection*{Outline}
In Section \ref{sec:classification}, we give a basic classification of BiB games to provide an overview of the research, followed by some applications in Section \ref{sec:applications}. Section \ref{sec:model} contains a detailed definition of the model we use, which can be classified as a sequential, finite and dynamic process with random deletions. In Section \ref{sec:multiple-choice} we provide a brief definition of three different classes of random placement strategies and present the two placement algorithms outlined in the introduction: Greedy (Section \ref{sec:greedy}) and Always-go-left (Section \ref{sec:AlwaysGoLeft}). Section \ref{sec:analysis} contains a detailed analysis of both strategies for the case $m=n$.

\section{Classification of BiB games}
\label{sec:classification}
Several different variations of this problem exist, which may be classified as follows:
\begin{compactitem}
\item In a \emph{sequential} process, the balls arrive one after another and a ball is dealt with only after the previous one has been placed. Conversely, in a \emph{parallel} process, balls arrive in sets and the balls in one set are dealt with simultaneously~\cite{ABS98}. 
\item In a \emph{static} problem, the balls are placed into the bins and no balls are ever removed, whereas in a \emph{dynamic} problem balls may also be deleted from the system. In the random deletions model, for each newly inserted ball another one is chosen uniformly and independently at random and removed from the system~\cite{ABKU99, MRS01}. More elaborate schemes exists: An oblivious adversary might specify a sequence of insertions and deletions of balls in advance, without any knowledge of their placement~\cite{CFM+98}. Adler et al.~\cite{ABS98} studied another approach in the context of the parallel BiB problem, where the balls are stored in FIFO-order and at every time step the first ball in each bin is deleted. 
\item A \emph{finite} process is analysed for a given time interval that is known beforehand, whereas for an \emph{infinite} one, placements and deletions are given for an infinite time span~\cite{ABS98}.
\item In the lightly loaded case, $m$ is in the order of $n$, whereas in the heavily loaded case, $m$ is arbitrarily large. W.l.o.g., we will assume that $m = n$ for the lightly loaded case. 
\end{compactitem}

@@

\section{BiB model}
\label{sec:model}
Although several different variants of the BiB problem exist, we will focus on a basic model to highlight the crucial ideas in the placement strategies and their analyses. They are based on a sequential, infinite and dynamic process that uses an oblivious adversary, which provides a sequence $\sigma = \sigma_1 \sigma_2 \dots$, where $\sigma_t = (d_t, a_t)$ represents a request for a ball $d_t$ to be deleted and a different ball $a_t$ to be added to the bins at time step $t$. In our model, the ball $d_t$ associated with some request $\sigma_t$ is deleted at the beginning of time step $t$ and the ball $a_t$ has to be placed into one of the $n$ bins, without any knowledge of the following requests $\sigma_{t+1}$, $\sigma_{t+2},\dots$ This infinite sequence of insertions and deletions is fixed in advance and is thus independent of previously allocated balls and any decisions made by the allocation algorithm. We assume that, at any time $t$, at most $m = n$ balls are in the bins.

\section{Multiple-choice BiB}
\label{sec:multiple-choice}
The single-choice BiB strategy (Section \ref{sec:single-choice}) chooses one bin uniformly and independently at random for each ball. A multiple-choice BiB strategy generalizes this idea by  choosing $d \geq 2$ bins and placing the ball into the least full. We distinguish placement algorithms depending on how the $d$ locations (i.e., bins) are sampled~\cite{VOC03}.

@@

\subsection{The Greedy scheme}
\label{sec:greedy}
Consider the Greedy scheme: For each ball, choose $d \geq 2$ bins independently and uniformly at random and place the ball into the least full bin. If there are multiple bins with the same minimal load, choose one among them arbitrarily. Note that the $d$ selected bins are not necessarily distinct.

\begin{theorem}
\label{theorem:greedy}
Consider any sequence of insertions and deletions s.t.~at most $n$ balls are in the bins at any time. If the balls are placed into bins by the Greedy algorithm, then the maximum load is at most
\begin{align*}
\load \leq \frac{\ln \ln n}{\ln d}+ O(1) \quad ,
\end{align*}
w.h.p., for all $t$~\cite{ABKU99, BCSV06}.
\end{theorem}
This result is remarkable in several ways:
\begin{compactitem}
\item The apparently small change made for the Greedy algorithm compared to the single-choice BiB scheme results in an exponential decrease of the additive gap, even for $d=2$. 
\item Each additional choice ($d > 2$) decreases the gap by only a constant factor~\cite{MRS01}.
\end{compactitem}

@@

\begin{theorem}
\label{theorem:aglm}
Consider any sequence of insertions and deletions s.t.~at most $n$ balls are in the bins at any time. If the balls are placed into bins by the Always-go-left algorithm, the maximum load is at most 
\begin{align*}
\load \leq \frac{\ln\ln n}{d  \ln \Phi_d} + O(1) \quad ,
\end{align*}
w.h.p.,~for all $t$~\cite{VOC03}.
\end{theorem}
\noindent
V\"ocking~\cite{VOC03} also proved a matching lower bound for the lightly loaded case:
\begin{theorem}
\label{theorem:aglmLower}
For $m=n$, using an arbitrary allocation scheme that chooses $d$ bins for each ball at random (i.e., any allocation scheme in class 3) the maximum load is
\begin{align*}
\load = \frac{\ln \ln n}{d \ln \Phi_d} - O(1) \quad 
\end{align*}
\end{theorem}
\noindent
w.h.p.,~for all $t$~\cite{VOC03}. Berenbrink et al.~\cite{BCSV06} were able to show that the Always-go-left strategy is optimal for the heavily loaded case as well. We note that the combination of asymmetric tie-breaking and and non-uniform (but independent) sampling of the $d$ bins is crucial to achieve the bounds given in Theorems~\ref{theorem:aglm} and~\ref{theorem:aglmLower}.

@@

We will prove the bounds for $m=n$ for both the Greedy and the Always-go-left scheme in Sections~\ref{sec:analysis:greedy} and~\ref{sec:analysis:alg}, respectively. However, we will adopt a simplifying assumption to provide a clearer presentation of the proof technique. Section~\ref{sec:analysis:nondistinctBalls} will extend these proofs to remove this assumption.

We will use a proof technique called witness trees, which allows for a clean analysis of both problems and is applicable to a wide variety of BiB scenarios. Other proof techniques such as layered induction and a differential equations approach have been used in the literature; see, e.g.,~\cite{MRS01} for an overview of different proof techniques for the BiB problem. The proof for arbitrarily large $m$ is more involved and we refer the interested reader to~\cite{BCSV06, TW13}.

@@

\subsubsection{Symmetric witness trees}
\label{sec:analysis:definitionSymWT}
A symmetric witness tree (SWT) of order $L$ is a complete $d$-ary tree with $l=d^{L}$ leaves, where each node $v$ references a ball $b_v = \mathrm{ball}(v)$ and the nodes and edges of the SWT represent events that may occur or not. In general, the same ball may be referenced by several nodes. However, initially we will make use of the following assumption in the analysis.
\begin{assumption}
\label{assumption:independence}
All the events represented by the nodes and edges of a witness tree are stochastically independent. 
\end{assumption} 
\noindent
This implies that no ball is referenced by more than one node of a witness tree. A correct analysis will have to consider balls assigned to more than one node, but this simplifying assumption is useful to highlight the differences between the symmetric and asymmetric allocation schemes. Section~\ref{sec:analysis:nondistinctBalls} will extend the analysis to witness trees with redundant balls.

We define edge and leaf events:
\begin{description}
\item [Edge event:] Consider some edge $e = (u,v)$ of the witness tree, s.t.~$v$ is the $i$-th child of $u$. The edge $e$ represents the event that the $i$-th location of the ball $b_u$ points to the same bin as one of the locations of the ball $b_v$. 
\item [Leaf event:] A leaf node $v$ of the witness tree represents the event that each of the $d$ locations of the ball $b_v$ points to a bin that contains at least three additional balls (i.e., balls that are not represented by a node of the tree), at the time of the insertion of the ball $b_v$.
\end{description}
To avoid dependencies, the crucial step is to define edge events in terms of the possible locations of the balls, not by the bin that a ball is finally assigned to.

@@

\subsubsection{Probability of activation}
\label{sec:analysis:probabilitySymWT}
In this section, we derive an upper bound for the probability that some SWT of order $L$ is activated.

@@

The probability that a leaf event is activated is bounded from above by $3^{-d}$: By the definition of leaf events, each of the $d$ locations of a ball $b_v$, where $v$ is a leaf node, has to point to a bin with at least 3 additional balls in it. The number of those bins is bounded by $\lfloor n/3 \rfloor$ at all time steps $t$, implying a probability of at most $1/3$ that one such bin is chosen. This implies a probability of at most $3^{-d}$ that all the $d$ bins are chosen from among those with at least 3 balls in them. Thus, the probability that all leaf events are activated is at most $p_\mathrm{l} = 3^{-d l}$, where $l = d^{L}$ is the number of leaves in an SWT.

Thus, the probability that an activated SWT exists at time $t$ is at most 
\begin{align*}
P_\mathrm{s} &\leq w  p_\mathrm{e}  p_\mathrm{l} 
		= n^{k}  \left(\frac{d}{n}\right)^{k-1}  3^{-d  l} 
		= n  d^{k-1}  3^{-d  l}\quad .
\end{align*}
Using $k \leq 2l$, we get
\begin{align*}
P_\mathrm{s} &\leq n  d^{2 l}  3^{-d  l} 
		= n  \left(d^2\right)^l  \left( 3^{-d} \right)^l\quad ,
\end{align*}
which we simplify by using $2d^2 \leq 3^d$ to
\begin{align*}
P_\mathrm{s} &\leq n  \left(\frac{3^d}{2}\right)^l  \left( 3^{-d}\right)^l 
		= n  2^{-l}  3^{d l}  3^{-d  l} 
		= n 2^{-l}		
		= n 2^{-d^L} \quad .		
\end{align*}
For any $\alpha > 0 $, let
\begin{align*}
L &\geq \log_d\log_2 n + \log_d\left(1+\alpha\right) \neqn{swt:chooseL} \quad .
\end{align*}
Then, 
\begin{align*}
P_\mathrm{s} &\leq n  2^{{-d}^{\log_d\log_2 n + \log_d\left(1+\alpha\right)}} 
		= n  2^{-{d}^{\log_d\log_2 n} d^{\log_d \left(1+\alpha\right)}} \\
		&= n  \left( 2^{-\log_2 n  \left(1+ \alpha \right)} \right) 
		= n  n^{-\left(1+\alpha\right)} = n^{-\alpha} \neqn{swt:result}\quad .
\end{align*}
Thus, for any $\alpha>0$ and large enough $L$ as specified in Eq.~\ref{eqn:swt:chooseL} the probability that an SWT of order $L$ is activated is bounded from above as $P_\mathrm{s}\leq n^{-\alpha}$.
\end{proof}

Since the probability for the activation of an SWT of order $L$ upper-bounds the probability of the bad event that some bin contains more than $L+3$ balls, Eq.~\ref{eqn:swt:result} is also an upper bound for the latter. Note that Eq.~\ref{eqn:swt:chooseL} can be rewritten as
\begin{align*}
L &\geq  \frac{\ln\log_2 n}{\ln d} + \frac{\ln(1+\alpha)}{\ln d} 
   = \frac{\ln \log_2 n}{\ln d} +O(1) 
  \geq \frac{\ln \ln n}{\ln d} +O(1) \neqn{swt:boundL} \quad ,
\end{align*}
thus proving Theorem~\ref{theorem:greedy}, under Assumption~\ref{assumption:independence}. \qed

@@

\subsubsection{Asymmetric witness trees}
\label{sec:analysis:definitionAsymWT}
An AWT differes from an SWT only by its structure, which is a $d$-step Fibonacci tree $T_d\left(k \right)$:

\begin{compactitem}
\item $T_d(1)$ is a single node and  $T_d(2) = T_d(1)$.
\item For $3\leq k \leq d$, $T_d(k)$ is a tree with $k-1$ children, which are the roots of the trees $T_d(k-1),\ldots,T_d(1)$.
\item For $k>d$, $T_d(k)$ is a tree with $d$ children, which are the roots of the trees $T_d(k-1),\ldots,T_d(k-d)$.
\end{compactitem}
A $d$-ary AWT of order $L$ has the structure of a Fibonacci tree $T_d(d  L+1)$ and thus has $l=F_d(d L + 1) \geq \Phi_d^{d L-1}$ leaves (Eq.~\ref{eqn:capocelli}).
Furthermore, for each node we define a label $(h, i)$, where $h, i \in \mathbb{N}$, $0 \leq h \leq L$ and  $1\leq i \leq d$:
\begin{compactitem}
\item The root nodes of the Fibonacci trees $T_d(1)$ and $T_d(2)$ are assigned the labels $(0,1)$ and $(0,2)$, respectively. Thus, nodes labelled $(0,1)$ or $(0,2)$ do not have any children.
\item A node with label $(0, i)$, where $i>2$, has $i-1$ children with the labels $(0,i-1),\dots,(0,1)$.
\item A node with label $(h,i)$, where $h>0$, has $d$ children labelled $(h,i-1),\dots,(h,1)$,$(h-1,d),\dots,(h-1, i)$.
\item The root of a Fibonacci tree $T_d(d h+i)$ is labelled $(h, i)$. 
\end{compactitem}
Thus, the root node of an AWT of order $L$ is assigned the label $(L, 1)$. Note that the labels are unique for the set of children of any node $v$, but not necessarily for all nodes of the tree. 
\begin{lemma}
When using the Always-go-left algorithm to put balls into bins, a bin containing more than $L+3$ balls activates an AWT of order $L$.
\end{lemma}
\begin{proof}
The following invariant will be maintained while assigning balls to the nodes of the AWT:
\begin{invariant}
A ball $b_v$ referenced by a node $v$ with label $(h,i)$ was placed in a bin $x$ belonging to group $i$ that, at the time of insertion of the ball $b_v$, contained at least $h+3$ other balls. The ball represented by the root node may be put in a bin belonging to any group.
\end{invariant}
\noindent
Let $x$ be the bin that contains $L+4$ balls and assign the topmost ball $b_t$ in $x$ to the root node of the AWT. Thus, at the time of insertion of $b_t$, each of the $d$ locations of this ball points to a bin containing at least $L+3$ balls. These are the $d$ bins that were sampled by the Always-go-left scheme, i.e., bin $i$ belongs to the $i$-th group of bins, where each group consists of $n/d$ bins. The child node of the root with label $(L-1, i)$ is assigned the topmost ball of the $i$-th location. The assignment of balls to nodes then proceeds recursively: Given a node $v$ labelled $(h, i)$ that references a ball $b_v$ (which was put into some bin of the $i$-th group that, at the time of insertion of ball $b_v$, contained at least $h+3$ other balls), we can conclude the following:
\begin{compactitem}
\item For $1\leq j < i$, the $j$-th location of the ball $b_v$ contains at least $h+4$ balls, at the insertion time of the ball $b_v$. If it contained less balls than the $i$-th location, the Always-go-left scheme would have put the ball $b_v$ into the bin pointed to by location $j$. The topmost ball in location $j$ is assigned to the child $u$ of $v$ with label $(h,j)$. 
\item For $i < j \leq d$, the $j$-th location contains at least $h+3$ balls. If it contained less, the Always-go-left scheme would have put the ball $b_v$ into the bin location $j$ points to. The topmost ball in the bin that location $j$ points to is assigned to the child node $u$ of $v$ labelled with $(h-1, j)$.
